From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-10 upi.tex
slides

The early developments of reversible programming languages started with a
conventional programming language, e.g., an extended $\lambda$-calculus, and either
\begin{enumerate*}[(i)]
\item extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
\item imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
\end{enumerate*}
More modern approaches recognize that reversible programming languages require
a fresh approach and should be designed from first principles without the
detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.

@@

By restricting to finite types, two operational semantics emerge:
\begin{enumerate*}{(i)}
\item a (reversible) operational semantics of programs over finite types, and
\item a (reversible) operational semantics of program transformers of such programs.
\end{enumerate*}
Such programs can be interpreted as generalized permutations, as they work on
structure as well as the underlying skeleton of finite types, aka the natural numbers.
Many of the transformers can be \emph{oriented}, and thus be interpreted as
\emph{optimizations} in some sense.

@@

We will consider a general statement of \emph{univalence} for type families. We
start with a some preliminaries. Fig.~\ref{fig:fib} depicts a type family
$P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space $P(x)$. As
everything is continuous, we expect this type family to respect identities. In
other words if $x:A$ and $y:A$ are identified using a path $x \equiv y$, it must
be the case that the spaces $P(x)$ and $P(y)$ become equivalent
$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is however no
reason to expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be
themselves equivalent, and indeed we will provide a few examples in which they
are not. When they are indeed equivalent, the fibration is a \emph{univalent
  fibration}.

\begin{figure}
\begin{tabular}{c@{\qquad\qquad}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.5,-2) {Fiber $P(x)$};
  \draw (1,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.5,-2) {Fiber $P(y)$};
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.8,-2) {Fiber $P(x)$};
  \draw (1.3,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.8,-2) {Fiber $P(y)$};
  \draw[below,cyan,thick] (-1,-5) -- (1,-5);
  \node[below,cyan,thick] at (0,-5) {$\equiv$};
  \draw[->,red,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
  \draw[->,red,ultra thick] (0.8,-2.3) to [out=-135, in=-45] (-0.8,-2.3);
  \node[red,ultra thick] at (0,-2) {$\simeq$};
\end{tikzpicture}
\end{tabular}
\caption{\label{fig:fib}(left)Type family $P : A \rightarrow \mathcal{U}$ as a
  fibration with total space $\Sigma_{(x:A)} P(x)$; (right) a path $x \equiv y$
  induces an equivalence between the spaces $P(x)$ and $P(y)$}
\end{figure}

@@

\medskip\noindent The identity function and boolean negation induce two distinct
equivalences in the space $P(\star) = \mathbb{2} \simeq \mathbb{2} =
P(\star)$. As there is only one path $\star \equiv star$ in the base space, the
fibration $P$ is not univalent. Similarly, there is one equivalence
$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but this certainly does not
induce a path $\fc \equiv \tc$. As the examples illustrate there is a delicate
balance required for the fibration to be univalent; otherwise some distinct
equivalences would be identified (like for $P$) or distinct points in the base
space would be identified (like for $Q$).

>>>>

The early developments of reversible programming languages started with a
conventional programming language, e.g., an extended $\lambda$-calculus, and either
\begin{enumerate}
\item extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
\item imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
\end{enumerate}
More modern approaches recognize that reversible programming languages require
a fresh approach and should be designed from first principles without the
detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.

@@

By restricting to finite types, two operational semantics emerge:
\begin{enumerate}
\item a (reversible) operational semantics of programs over finite types, and
\item a (reversible) operational semantics of program transformers of such programs.
\end{enumerate}
Such programs can be interpreted as generalized permutations, as they work on
structure as well as the underlying skeleton of finite types, aka the natural numbers.
Many of the transformers can be \emph{oriented}, and thus be interpreted as
\emph{optimizations} in some sense.

@@

We will consider a general statement of \emph{univalence} for type families. We
start with a some preliminaries. Fig.~\ref{fig:fib} depicts a type family
$P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space $P(x)$. As
everything is continuous, we expect this type family to respect identities. In
other words if $x:A$ and $y:A$ are identified using a path $x \equiv y$, it must
be the case that the spaces $P(x)$ and $P(y)$ become equivalent
$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is no reason to
expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be themselves
equivalent for all $x$ and $y$. And indeed they are most frequently not
equivalent. In the rare cases in which they are equivalent, the fibration is a
\emph{univalent fibration}.

\begin{figure}
\begin{tabular}{c@{\qquad\qquad}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.5,-2) {Fiber $P(x)$};
  \draw (1,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.5,-2) {Fiber $P(y)$};
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.8,-2) {Fiber $P(x)$};
  \draw (1.3,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.8,-2) {Fiber $P(y)$};
  \draw[below,cyan,thick] (-1,-5) -- (1,-5);
  \node[below,cyan,thick] at (0,-5) {$\equiv$};
  \draw[->,red,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
  \draw[->,red,ultra thick] (0.8,-2.3) to [out=-135, in=-45] (-0.8,-2.3);
  \node[red,ultra thick] at (0,-2) {$\simeq$};
\end{tikzpicture}
\end{tabular}
\caption{\label{fig:fib}(left) Type family $P : A \rightarrow \mathcal{U}$ as a
  fibration with total space $\Sigma_{(x:A)} P(x)$; (right) a path $x \equiv y$
  in the base space induces an equivalence between the spaces $P(x)$ and $P(y)$}
\end{figure}

@@

\medskip\noindent The identity function and boolean negation induce two distinct
equivalences in the space $P(\star) = \mathbb{2} \simeq \mathbb{2} =
P(\star)$. As there is only one path $\star \equiv \star$ in the base space, the
fibration $P$ is not univalent. Similarly, there is one equivalence
$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but this certainly does not
induce a path $\fc \equiv \tc$. As the examples illustrate there is a delicate
balance required for the fibration to be univalent; otherwise some distinct
equivalences would be identified (like for $P$) or distinct points in the base
space would be identified (like for $Q$).
